---
title: "Using SNOMED dictionaries and codelists"
author: "Anoop Shah"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using SNOMED dictionaries and codelists}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Using SNOMED dictionaries and codelists
=======================================

This package is designed to make it easier to use SNOMED CT in R, including searching the SNOMED CT dictionary and relations, and creating and manipulating SNOMED CT codelists.

Basic introduction to SNOMED CT
-------------------------------

SNOMED CT is a clinical terminology system which contains thousands of concepts, each of which has a distinct meaning. Each concept has a unique concept ID, and one or more descriptions (synonyms). It also contains a knowledge model (ontology), specifying which concepts are subtypes of another concept or associated in other ways.

The SNOMED dictionaries are contained within four key tables:

- CONCEPT - one row per concept, with key metadata about the concepts
- DESCRIPTION - one row per description
- STATEDRELATIONSHIP - directly stated relationships
- RELATIONSHIP - relationships between terms inferred from other relationships

Each SNOMED CT concept may have any number of synonyms, but there are the following special types:

- Fully Specified Name - this is the most precise description of the concept, but is too verbose for general use.
- Preferred - these is the synonym that is preferred for clinical display in a particular language, and is recorded in the language-specific reference set

Each concept also has a semantic tag, which denotes what type of concept it is (e.g. 'disorder' or 'organism'). This is currently recorded in parentheses at the end of the Fully Specified Name, and can be extracted using the semanticType function in this package.

Loading the SNOMED CT dictionaries
----------------------------------

This package contains functions to import a set of SNOMED CT release files from NHS Digital TRUD (https://isd.digital.nhs.uk/trud3/user/guest/group/0/home). The International and UK release files are provided in separate downloads, and need to be combined to create the whole dictionary. This can be done using the loadSNOMED() function and specifying a vector of folder paths to load.

For most users the 'Snapshot' files, containing the current versions of the entries, will be the most useful, and the sample dictionaries in this package are based on the Snapshot file format. The 'Delta' files contain changes since the previous version, and the 'Full' files contain a history of the changes to each entry.

When the NHS Digital SNOMED CT files are extracted, they will be in a folder such as 'SnomedCT_InternationalRF2_PRODUCTION_20200731T120000Z', and the relevant tables will be in the subfolder 'Snapshot/Terminology'. Code such as the following can be used to load the tables into R (the file paths should be changed to the actual file paths on your system).

```
# Load package
require(Rdiagnosislist)

# Load UK and International SNOMED CT release files into an
# R environment called 'SNOMED'
SNOMED <- loadSNOMED(c(
  'SnomedCT_InternationalRF2_PRODUCTION_20200731T120000Z/Snapshot/Terminology',
  'SnomedCT_UKClinicalRF2_PRODUCTION_20210317T000001Z/Snapshot/Terminology'))

# Save the 'SNOMED' environment to a file on disk
saveRDS(SNOMED, file = 'mySNOMED.RDS')

# Reload the 'SNOMED' environment from file
SNOMED <- readRDS('mySNOMED.RDS')
```

Using R environments
--------------------

The SNOMED CT dictionary files are loaded into an R 'environment', which is an object that can contain other objects. This is a convenient way in R to store a group of objects without cluttering up the global environment with too many individual objects. It also allows different versions of SNOMED CT dictionaries to be used side by side, by loading them into different environments. For ease of use, many of the functions in the package will search for the dictionaries in an environment named 'SNOMED' by default, but for programming use it is recommended to specify the environment explicitly in function calls to avoid unexpected errors. 

For the purpose of this vignette, we will create a sample set of SNOMED CT files from the sample dictionaries included with the package. 

```{r}
# The sampleSNOMED() function returns an environment containing
# the sample dictionaries
require(Rdiagnosislist)
TEST <- sampleSNOMED()

# TEST is now an environment containing the sample SNOMED CT dictionary.
# Objects within the environment can be retrieved using the $ operator
# or the 'get' function. We will export the sample dictionaries to a
# temporary folder to show how to reload them using loadSNOMED()

for (table in c('Concept', 'Description', 'Relationship',
  'StatedRelationship')){
  write.table(get(toupper(table), envir = TEST), paste0(tempdir(),
    '/sct_', table, '_test.txt'), row.names = FALSE, sep = '\t', quote = FALSE)
}

# loadSNOMED searches for files containing _Concept_, _Description_,
# _Relationship_ or _StatedRelationship_, as in the actual SNOMED CT
# release files.

# Import using the loadSNOMED function
SNOMED <- loadSNOMED(tempdir(), active_only = FALSE)
```

SNOMED CT concepts IDs in R
---------------------------

SNOMED CT concept IDs are long integers which need to be represented using the integer64 data type in R. This is not available in base R but is provided in the bit64 package which is automatically loaded with this package. They must not be stored as short integer or numeric values because they cannot be stored precisely and may be incorrect.

This package makes it easier to use SNOMED CT concept IDs because they can be supplied as character vectors and converted to 'SNOMEDconcept' vectors. The SNOMEDconcept class is a 64-bit integer class which can faithfully store SNOMED CT concept IDs, and is more memory-efficient than storing them as character vectors. If the SNOMED dictionary is available in the R environment, the concepts are displayed with their description in the default print method.

The function 'as.SNOMEDconcept' can be used to retrieve the SNOMED concept ID matching a description. This function also converts SNOMED CT concept IDs in other formats (numeric, 64-bit integer or character) into SNOMEDconcept objects.

```{r}
# Make sure the SNOMED environment is available and contains the SNOMED dictionary
as.SNOMEDconcept('Heart failure', SNOMED = SNOMED)

# To use the sample SNOMED dictionary for testing
as.SNOMEDconcept('Heart failure', SNOMED = sampleSNOMED())

# If an object named SNOMED containing the SNOMED dictionary is available
# in the current environment, it does not need to be stated in the
# function call
SNOMED <- sampleSNOMED()
as.SNOMEDconcept('Heart failure')

# The argument 'exact' can be used to specify whether a regular expression
# search should be done, e.g.
as.SNOMEDconcept('Heart f', exact = FALSE)

# The 'description' function can be used to return the descriptions of
# the concepts found. It returns a data.table with the fully specified 
# name for each term.
description(as.SNOMEDconcept('Heart f', exact = FALSE))

# The 'semantic type' function returns the semantic type of the concept
# from the Fully Specified Name
semanticType(as.SNOMEDconcept('Heart failure'))

# Functions which expect a SNOMEDconcept object, such as semanticType,
# will automatically convert their argument to SNOMEDconcept using the
# function as.SNOMEDconcept
semanticType('Heart failure')
```

Set operations using SNOMEDconcept
----------------------------------

This package provides versions of the set functions 'setdiff', 'intersect' and 'union' which work with SNOMEDconcept objects.

```{r}
# A list of concepts with a description containing the term 'heart'
# (not that all synonyms are searched, not just the Fully Specified Names)
heart <- as.SNOMEDconcept('Heart|heart', exact = FALSE, SNOMED = sampleSNOMED())

# A list of concepts containing the term 'fail'
fail <- as.SNOMEDconcept('Fail|fail', exact = FALSE, SNOMED = sampleSNOMED())

# Concepts with heart and fail
intersect(heart, fail)

# Concepts with heart and not fail
setdiff(heart, fail)

# Concepts with heart or fail
union(heart, fail)
```

These set operations can be used to create lists of SNOMED CT concepts of interest, similar to the way researchers use Read codes. However, SNOMED CT also allows the use of hierarchies and relationships to locate concepts by their meaning.

Using relationships between SNOMED CT concepts
----------------------------------------------

The most important relationship is the 'Is a' relationship, also known as parent and child. This makes it easy to find all specific concepts that are a subtype of a more general concept. The relationship functions (parents, ancestors, children and descendants) all take a SNOMEDconcept object as input, or attempt to convert their argument to SNOMEDconcept.

```{r}
SNOMED <- sampleSNOMED()

# Parents (immediate ancestors)
parents('Acute heart failure')

# Ancestors
ancestors('Acute heart failure')

# Children (immediate descendants)
children('Acute heart failure')

# Descendants
descendants('Acute heart failure')
```

Attributes of SNOMED CT concepts
--------------------------------

The 'hasAttributes' function can be used to find terms with particular attributes. For example, to find all disorders with a finding site of the heart, we can use the relatedConcepts function, which retrieves relationships from the RELATIONSHIP and STATEDRELATIONSHIP tables. 

In order to find the finding site of a disorder, we use the 'forward' relationship. In order to find disorders with a particular finding site, we use the relationship in the 'reverse' direction.

```{r}
require(Rdiagnosislist)
SNOMED <- sampleSNOMED()

# List all the attributes of a concept
print(attrConcept('Heart failure'))

# 'Finding site' of a particular disorder
relatedConcepts('Heart failure', 'Finding site')

# Disorders with a 'Finding site' of 'Heart'
relatedConcepts('Heart', 'Finding site', reverse = TRUE)
```

SNOMED CT codelists
-------------------

The SNOMED CT codelist data type allows a list of SNOMED CT concepts to be curated. It allows codelists to be expressed in a contracted (simplified) form with descendants implicitly included.

```{r}
SNOMED <- sampleSNOMED()

# Create a codelist containing all the descendants of
# the concept 'Heart failure'
my_heart_failure_codelist <- SNOMEDcodelist(
  as.SNOMEDconcept('Heart failure'), include_desc = TRUE)

# Original codelist
print(my_heart_failure_codelist)

# Expanded codelist
expanded <- expandSNOMED(my_heart_failure_codelist)
print(expanded)

# Contract codelist
contracted <- contractSNOMED(expanded)
print(contracted)

# Write out codelist to file
write.csv(my_heart_failure_codelist, file = paste0(tempdir(),
  '/hf_codes.csv'), row.names = FALSE)

# Reload codelist from file
reloaded_codelist <- as.SNOMEDcodelist(
  data.table::fread(paste0(tempdir(), '/hf_codes.csv')))
print(reloaded_codelist)
```

More information
----------------

For more information about SNOMED CT, visit the SNOMED CT international website: <https://www.snomed.org/>

SNOMED CT (UK edition) can be downloaded from the NHS Digital site: <https://isd.digital.nhs.uk/trud3/user/guest/group/0/home>

The NHS Digital terminology browser can be used to search for terms interactively: <https://termbrowser.nhs.uk/>

